Tracing the Hercules stream around the Galaxy

### 2010-09-09 tex/response.txt

Dear Editor,

Please find below the response to the referees' comments for the
submission "Tracing the Hercules Stream around the Galaxy" by myself
(MS# ApJ80617).

I have made changes following the referee's suggestions ...

A detailed list of the main changes is:

Please contact me by the information given below with any questions
regarding this re-submission.

With kind regards,

Jo Bovy

-
Jo Bovy
Ctr. for Cosmology & Particle Physics, 
Dept. of Physics
New York University
4 Washington Place
New York, NY 10003
jb2777@nyu.edu

In what follows the referee's comments are indented.

==================================================================

	Referee Report
	 Reviewer's Comments:

	 The paper titled Tracing the Hercules stream around the Galaxy by Jo
	 Bovy is concerned with the origin and detectability of this
	 elusive stream. Overall the paper is well written and
	 clear. I have a few major concerns and some minor concerns
	 that should be addressed before the paper can be accepted for
	 publication.

	 Major concerns:

	 1. Sect 4, page 6, last paragraph: The author discusses in
	 some detail, and shows plots to illustrate, the possible
	 implications of e.g.  errors in the distance determination
	 etc. The one error that is not address explicitly in the text
	 or plots is the error on the measured radial velocities - how
	 well do we need to determine these velocities in order to be
	 able to unambiguously detect the Hercules stream? 1, 5, or 10
	 km/s? Or some other number? And what happens when V0 is
	 absolutely known or if it also has an error? Perhaps the
	 result is that the errors do no matter, but in that case this
	 should be shown and arguments to why it does not matter
	 should be given. Also, will either the RVS on board Gaia or
	 APOGEE deliver radial velocities with small enough errors to
	 detect the Hercules stream?

	 2. A brief discussion about which tracers, e.g. dwarf stars,
	 red clump, red giant stars, to be used to test the model
	 should also be included. For example, is it necessary to
	 reach as far away from the solar neighbourhood as possible
	 (eg giants), or is the major test to get as high volume
	 density of tracers as possible (eg cool dwarfs). Such a
	 discussion would be highly useful for anyone wanting to
	 construct surveys for the very purpose of detecting
	 Hercules. This discussion does not need to be long.

	 3. I find Discussion in Sect 5 (for both Gaia and APOGEE)
	 somewhat sketchy. For example, no mention of the fact that
	 Gaia will provide radial velocities for stars down to roughly
	 17 magnitude is mentioned.  Should the reader conclude that
	 Hercules could be detected with the tangential velocities
	 alone? In that case, how does that work with the tests
	 presented in this article? Again, here the intrinsic
	 brightness of the tracers is of some interest.

	 4. The discussion could in general be deeper, and also connect to
	 the abstract (see below).

	 Additional concerns:

	 1. On page 2, last paragraph of the introduction it is stated
	 that "propose and ... test of the bar resonance model for the
	 Hercules stream". However, this statement is not really
	 carried through in the text. It might be a good idea to
	 connect back to this statement in e.g. Sect 5 and tell the
	 reader how "testable" this model is once all the
	 uncertainties of data and shortcomings of the missions has
	 been discussed. The test and properties of the bar are also
	 mentioned in the abstract, however, there is no discussion in
	 the paper of how the properties of the bar are constrained by
	 the proposed observational test. Add this discussion, which
	 would be very interesting.

	 2. Page 2, statement at the end of the first paragraph "..up
	 to the task of separating out these effects." This statement
	 should be supported with a few relevant references.

	 3. On page 5, sect 4, third paragraph, last sentence. I do
	 not understand what the last sentence means - could the
	 author please re-phrase it? The sentence somehow, does not
	 connect to the previous text.

	 Minor items:

	 1. Page 2, sect 2, end of the last sentence. "evaluated at
	  the initial orbit".  This is, for a person with English as a
	  second language, really hard to understand. It is not the
	  words that are hard but how it follows from the first part
	  of the sentence. For better clarity to all readers I suggest
	  that the author re-writes the last sentence, perhaps into
	  two.

This statement was just supposed to state the collisionless Boltzmann
equation, but I wrote "initial orbit" rather than "initial position
and velocity", since the phases or angles of the initial orbit are
irrelevant. This is because the initial distribution function is
time-independent, and thus only a function of the integrals of the
motion. I changed the wording to "initial position and velocity" but
added a comment about the time-indepence of the DF.

	  2. Page 4, sect 3, first paragraph. Hipparcos mission (at
	   the end of the first paragraph) should be
	   referenced. Perryman et al. 1997, A&A, 323, L49, and/or the
	   catalogue references.

I have added the catalog reference.

	   3. Page 5, sect 4, end of third paragraph. The last
	    sentence is unclear.  APOGEE is able to probe "the other
	    side of the Galaxy" with reference to which side of the
	    Galaxy should the "other" be taken?  Other to 90 - 270
	    degrees? A more explicit statement would be helpful to the
	    reader.

The "other" side of the Galaxy is defined two sentences before this
sentence as the region 90 < \phi < 270, where \phi is the
Galactocentric azimuth. I don't think it is necessary to repeat the
definition two sentences later.

	    4. Page 7, sect 5.1, first sentence, Gaia should have a
	    reference. A relevant reference is Perryman, M. A. C., et
	    al., A&A 369, 339 (2001).

That's quite right. I have added this reference.

	    5. Page 8, sect 5.2, first paragraph
	      Line 6 from top: hercules --> Hercules
	      Line 9 from top: APOGEE could trace --> APOGEE could 
	         potentially trace
	      	 (his is more in line with the other comments in the article
		 that APOGEE is jut on the edge of the suitable region etc)
	      Line 4 from the bottom: 
	         chemical abundances --> elemental abundances

All fixed!

### 2010-09-09 tex/hercules.tex
groups-simulations: Gardner reference update
groups-simulations: groups paper reference
hercules: first round of response

To simulate the effect of the bar on stellar orbits near the Solar
circle, we follow the approach of \citet{dehnen00a}. For each fixed
position in the Galaxy, we evaluate the velocity distribution function
at values for the radial---toward the Galactic center---and
tangential---in the direction of Galactic rotation---velocity by
backward-integrating these velocities in the Galactic potential to
obtain the intial orbit before bar formation. We posit that at this
epoch the old stellar disk was in a steady-state that can be described
by a simple distribution function that is a function of the energy and
angular momentum of the orbit alone. It follows from the collisionless
Boltzmann equation that the value of the velocity distribution at the
present position and velocity is equal to that of the steady-state
distribution function before bar formation evaluated at the initial
orbit.

@@

\citet{dehnen00a} only simulated the velocity distribution at the
position of the Sun for various combinations of the model parameters
defined in the previous Section. Nevertheless, his results on how the
velocity distribution at the Sun changes with the angle between the
Sun-Galactic center line and the bar could also be interpreted as
showing the velocity distribution on the Solar circle for different
Galactocentric azimuths, fixing the bar angle. In this way, these
simulations can make predictions for what we might find when we study
the velocity distribution in regions beyond those probed by the
\hipparcos\ mission.

@@

The astrometric \Gaia\ mission will measure the parallaxes and proper
motions of up to one billion stars, most of which will be disk
stars. With parallax and proper motion accuracies down to 10
microarcseconds, \Gaia\ will be able to probe the kinematics of the
disk out to several kpc in all directions
\citep{bailerjones08a}. Crucially, it will provide large samples of
stars in many of the promising regions in
\figurename~\ref{fig:rphi2d}, which will be able to verify the
predictions laid out in this paper. This will also allow the dynamical
properties of the bar to be tightly constrained, something that will
be hard for \Gaia\ to do using direct observations of the bulge due to
the large extinction toward the Galactic center \citep{robin05a}.

@@

Since the extinction in the optical bandpasses is rather large in the
directions identified as promising in \figurename~\ref{fig:detect}
\citep[$A_V \approx 8$ mag toward $l \approx
270^{\circ}$;][]{marshall06a}, spectroscopic surveys in the
near-infrared such as \apogee\ are the best option for seeing the
Hercules feature in the line-of-sight direction. The range of Galactic
longitudes that \apogee\ can target along the disk is $-5^{\circ} \leq
l \leq 250^{\circ}$ (SDSS-III Project Description\footnote{Available
at \url{http://sdss3.org/collaboration/description.pdf}~.}). From
\figurename~\ref{fig:detect} this limit is right on the edge of where
we might hope to detect the Hercules feature. The hercules bump in the
direction $l = 250^{\circ}$ is about half as strong as in
\figurename~\ref{fig:1dvar}, but remains as a clear signature of the
bar. Since \apogee\ is expected to start taking data in 2011 and to
finish operations before the first \Gaia\ data release, \apogee\ could
trace the Hercules stream around the Galaxy before \Gaia\ does. Since
\apogee\ will also obtain detailed chemical abundances for its target
stars, this would also allow for a more detailed analysis of the
abundances of stars in the Hercules stream \citep[following,
\eg,][]{Bensby07a,Bovy10a}, which could shed more light on radial
mixing in the Galactic disk.

\acknowledgements It is a pleasure to thank Mike Blanton and David
W.~Hogg for helpful discussions and assistance.  Financial support for
this project was provided by the National Aeronautics and Space
Administration (grant NNX08AJ48G) and the National Science Foundation
(grant AST-0908357).

@@

\bibitem[Bovy \& Hogg(2010)]{Bovy10a} Bovy,~J.~\& Hogg,~D.~W., 2010,
  \apj, in press

@@

@@

\bibitem[Gardner \& Flynn(2010)]{gardner10a}
  Gardner,~E.~\& Flynn,~C., 2010,
  \mnras, in press

@@

>>>>

To simulate the effect of the bar on stellar orbits near the Solar
circle, we follow the approach of \citet{dehnen00a}. For each fixed
position in the Galaxy, we evaluate the velocity distribution function
at values for the radial---toward the Galactic center---and
tangential---in the direction of Galactic rotation---velocity by
backward-integrating these velocities in the Galactic potential to
obtain the intial orbit before bar formation. We posit that at this
epoch the old stellar disk was in a steady-state that can be described
by a simple distribution function that is a function of the energy and
angular momentum of the orbit alone. It follows from the collisionless
Boltzmann equation that the value of the velocity distribution at the
present position and velocity is equal to that of the steady-state
distribution function before bar formation evaluated at the initial
position and velocity---or, equivalently, at the initial orbit since
the initial distribution function is time-independent.

@@

\citet{dehnen00a} only simulated the velocity distribution at the
position of the Sun for various combinations of the model parameters
defined in the previous Section. Nevertheless, his results on how the
velocity distribution at the Sun changes with the angle between the
Sun-Galactic center line and the bar could also be interpreted as
showing the velocity distribution on the Solar circle for different
Galactocentric azimuths, fixing the bar angle. In this way, these
simulations can make predictions for what we might find when we study
the velocity distribution in regions beyond those probed by the
\hipparcos\ mission \citep{ESA97a}.

@@

The astrometric \Gaia\ mission will measure the parallaxes and proper
motions of up to one billion stars, most of which will be disk stars
\citep{Perryman01a}. With parallax and proper motion accuracies down
to 10 microarcseconds, \Gaia\ will be able to probe the kinematics of
the disk out to several kpc in all directions
\citep{bailerjones08a}. Crucially, it will provide large samples of
stars in many of the promising regions in
\figurename~\ref{fig:rphi2d}, which will be able to verify the
predictions laid out in this paper. This will also allow the dynamical
properties of the bar to be tightly constrained, something that will
be hard for \Gaia\ to do using direct observations of the bulge due to
the large extinction toward the Galactic center \citep{robin05a}.

@@

Since the extinction in the optical bandpasses is rather large in the
directions identified as promising in \figurename~\ref{fig:detect}
\citep[$A_V \approx 8$ mag toward $l \approx
  270^{\circ}$;][]{marshall06a}, spectroscopic surveys in the
near-infrared such as \apogee\ are the best option for seeing the
Hercules feature in the line-of-sight direction. The range of Galactic
longitudes that \apogee\ can target along the disk is $-5^{\circ} \leq
l \leq 250^{\circ}$ (SDSS-III Project Description\footnote{Available
  at \url{http://sdss3.org/collaboration/description.pdf}~.}). From
\figurename~\ref{fig:detect} this limit is right on the edge of where
we might hope to detect the Hercules feature. The Hercules bump in the
direction $l = 250^{\circ}$ is about half as strong as in
\figurename~\ref{fig:1dvar}, but remains as a clear signature of the
bar. Since \apogee\ is expected to start taking data in 2011 and to
finish operations before the first \Gaia\ data release, \apogee\ could
potentially trace the Hercules stream around the Galaxy before
\Gaia\ does. Since \apogee\ will also obtain detailed elemental
abundances for its target stars, this would also allow for a more
detailed analysis of the abundances of stars in the Hercules stream
\citep[following, \eg,][]{Bensby07a,Bovy10a}, which could shed more
light on radial mixing in the Galactic disk.

\acknowledgements It is a pleasure to thank the anonymous referee,
Mike Blanton, and David W.~Hogg for helpful discussions and
assistance.  Financial support for this project was provided by the
National Aeronautics and Space Administration (grant NNX08AJ48G) and
the National Science Foundation (grant AST-0908357).

@@

\bibitem[Bovy \& Hogg(2010)]{Bovy10a} Bovy,~J.~\& Hogg,~D.~W., 2010,
  \apj, 717, 617

@@

\bibitem[{{ESA}(1997)}]{ESA97a}
{ESA}, 1997, {The \emph{Hipparcos} and Tycho Catalogues} (ESA SP-1200; Noordwijk: ESA)

@@

\bibitem[Gardner \& Flynn(2010)]{gardner10a}
  Gardner,~E.~\& Flynn,~C., 2010,
  \mnras, 405, 545

@@

\bibitem[Perryman \etal(2001)]{Perryman01a}
  Perryman,~M.~A.~C., \etal, 2001,
  \aap, 369, 339
